Unifying intraspecific and interspecific variation in tropical tree mortality

### 2017-06-01 ms/supp.Rnw
Finished draft
- length of main text now within allowed 6 pages
- reworked final part of introduction, merging sections from the list of questions and methods into single para.
- shortened methods and added detail
- added intro to Supp Matt
- added details on numerical optimisations to SuppMatt

Our model is presented in detail in ...... Some intro text.

@@

To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}. This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual.

@@

\section{Mortality model}

We modelled tree mortality using survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$. As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

As we have multiple hypothesized functional forms of $\lambda_i$, we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases. The instantaneous rate of mortality is
\begin{equation}
\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, X_i \right)}^{\text{Growth dependent}} +
\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}\right) \times \overbrace{\delta_{t[i]}}^{\text{Census effect}}
\end{equation}  
In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species as a function of both wood density (measured at species level) and a species random effects, e.g.
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs. The form of sub-model in eq. \ref{eq:submodels} ensures that parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ remain positive. On a log scale this model equates to an additive linear model , i.e.
\begin{equation}\label{eq:submodels_log}
\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho.
\end{equation}

We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm$^{-3}$. This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm$^{-3}$. Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.

@@

>>>>

Here we provide some further details on methods and additional results, supplementing those presented in the main text. Section S1-Estimating True Growth describes a probabilistic methods used to estimate \textit{``true growth''} from field measurements,, taking into account measurement error and the distribution of growth rate across the community. Section S2-Parameter distributions describes the distributions applied to species-level random effects and priors put over parameters in the fitted model. Section S3-Numerical optimisation outlines some numerical concerns in estimating logloss, while Section S4-Example Stan code gives an example of the code fitted to the data. Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.

@@

 Growth rates were estimated from field measurements of diameter, which inevitably include observation error. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative. To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community. To achieve this we used another dataset collected at BCI to explicitly to examine observation errors\cite{Condit:2012nz}. This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual.

@@

\section{Parameter distributions}

@@

To evaluate eq. 5 from the main text (log loss), we apply the following substitutions, to reduce the number of $\log$ and $\exp$ operations, and thereby maintain numerical accuracy. From the main text, equation 5 states that the log loss for a single data point is
\begin{equation} \label{eq:logloss_i}
\mathcal{L}_i = S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3} = 1 - \exp\left(-\Lambda\right)$ is the probability an individual dies between times $t_2$ and $t_3$ (from eq. 1), $\Lambda$ is the cumulative hazard from $t_2\rightarrow t_3$, and $S_i$ is the observed outcome (0 = alive, 1 = died). Note that $\textrm{Pr}(S_i =0) = 1- p_{i,t_2\rightarrow t_3} = \exp\left(-\Lambda\right)$. Numerical calculations of log loss can then be simplified as follows:

If $S_i =0 \rightarrow \mathcal{L}_i = \log(1-p) = -\Lambda$.

If $S_i = 1  \rightarrow \mathcal{L}_i = \log(p) = \textrm{log1m\_exp}(-\Lambda)$.

\noindent In Stan, the function $ \textrm{log1m\_exp}$ is is defined so that $ \textrm{log1m\_exp}(x)$ has the same value as $\log(1.0-\exp(x))$, but the computation is faster, more accurate, and more stable \cite{StanDevelopmentTeam:2015uz}.
